On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-11-28 paper.tex
... better now

\begin{theorem}
\label{theo: funt, form}
For $a, b \in \mathbb R^+$, $a \geq b$ and $\len \in \mathbb R^+$,
\begin{equation}
\label{eq: theo: funt, form; funt}
\funt(\len) = \isoli+\jsoli-1
\end{equation}
with
\begin{align}
\label{eq: theo: funt, form; i}
\isoli &= \left\lceil \frac{a + b \Re \sqrt{4\len^2/(a^2+b^2)-1}}{2a} \right\rceil + 1, \\
\label{eq: theo: funt, form; j}
\jsoli &= \left\lceil \frac{\sqrt{\len^2-(\isoli-2)^2a^2}}{b} \right\rceil + 1.
\end{align}
The function $\funt$ is piecewise constant and left-continuous, with unit-height jumps. A jump occurs at $\len$ if and only if $\len = \funl(\tiles)$ for some $\tiles \in \mathbb N$, $\tiles \geq 4$; and then $\funt(\len) = \tiles-1$, $\lim_{h \rightarrow 0+} \funt(\len+h) = \tiles$.
\end{theorem}

@@

The pair $(\isolr, \jsolr)$ can be obtained as a solution of the equation system
\begin{align}
\label{eq: icont jcont 1}
(\isolr-2)^2 a^2 + (\jsolr-2)^2 b^2 &= \len^2, \\
\label{eq: icont jcont 2}
\jsolr = \frac{\isolr a^2}{b^2} - \frac{5a^2}{2b^2} + \frac 3 2.
\end{align}
Substituting \eqref{eq: icont jcont 2} into \eqref{eq: icont jcont 1} gives a quadratic equation for $\isolr$,
\begin{equation}
\label{eq: quadratic, i}
4a^2(\isolr-2)^2 - 4a^2(\isolr-2) + a^2+b^2-\frac{4\len^2 b^2}{a^2+b^2} = 0,
\end{equation}
which is solved as
\begin{equation}
\label{eq: icont jcont 3}
\isolr = \frac{a \pm b \sqrt{4\len^2/(a^2+b^2)-1}}{2a} + 2,
\end{equation}
and then $\jsolr$ can be obtained from \eqref{eq: icont jcont 2}. This can produce zero, one or two real-valued solution pairs $(\isolr, \jsolr)$, which respectively corresponds to the solid line in Figure~\ref{fig: ijLS_1p35_detail} being exterior, tangent or secant to the circle (the figure shows the latter case).

A solution $(\isolr, \jsolr)$ is of interest only if it is real-valued with $\isolr, \jsolr \geq 2$. This solution, if it exists, is always associated with the positive sign in \eqref{eq: icont jcont 3} (the negative sign either yields the same solution, in the tangent case, or yields a solution with $\jsolr < 2$, in the secant case). In this regard, three cases need to be distinguished: there are no real-valued solution pairs $(\isolr, \jsolr)$; there are one or two but do not satisfy $\isolr, \jsolr \geq 2$; there are one or two and and one of them satisfies that inequality.

If $\len < \sqrt{a^2+b^2}/2$, \eqref{eq: icont jcont 3} yields no real-valued solutions for $\isolr$. Since $a \geq b$, $\len < \sqrt{a^2+b^2}/2$ implies that $\len < a$, and therefore any achievable $(i,j)$ pair, if any, will have $i=2$. Thus in this case $\isoli$ should be set to $2$; and then the maximum achievable $\jsoli$ will be obtained from \eqref{eq: icont jcont 2}.

For $\len \geq \sqrt{a^2+b^2}/2$ there are either two real-valued solutions or one real-valued double solution for $\isolr$. These solutions will give $\jsolr < 2$ if $\isolr < (b^2/a^2+5)/2$, as can be seen setting $\jsolr=2$ in \eqref{eq: icont jcont 2}; or equivalently if $\sqrt{a^2+b^2}/2 \leq \len < (a^2+b^2) / (2a)$, as can be seen substituting $\jsolr < 2$ and $\isolr < (b^2/a^2+5)/2$ in \eqref{eq: icont jcont 1}. Thus for $\len$ in this range, since $\isolr < (b^2/a^2+5)/2 < 3$ for $a \geq b$, only pairs with $i=2$ are achievable again, and therefore $\isoli$ must also be $2$.

Lastly, for $\len \geq (a^2+b^2) / (2a)$ the expressions \eqref{eq: icont jcont 2} and \eqref{eq: icont jcont 3} with positive sign give $\isolr, \jsolr \geq 2$, and $\isoli$ should be taken as the greatest integer less than $\isolr$, i.e.~$\lceil \isolr \rceil-1$.

The three cases are unified by taking the real part of \eqref{eq: icont jcont 3} and computing $\isoli = \lceil \isolr \rceil-1$, as can be easily checked. This yield \eqref{eq: theo: funt, form; i}. Once $\isoli$ is known, $\jsoli$ is computed, according to \eqref{eq: theo: funt, form; j}, as the greatest integer such that $(\isoli a,\jsoli b)$ is within the circle determined by $\len$. This ensures that $(\isoli,\jsoli)$ is achievable with lengths less than $\len$.

>>>>

\begin{theorem}
\label{theo: funt, form}
For $a, b \in \mathbb R^+$, $a \geq b$ and $\len \in \mathbb R^+$,
\begin{equation}
\label{eq: theo: funt, form; funt}
\funt(\len) = \isoli+\jsoli-1
\end{equation}
with
\begin{align}
\label{eq: theo: funt, form; i}
\isoli &= \left\lceil \frac 3 2 + \frac b a \Re \sqrt{\frac{\len^2}{a^2+b^2}-\frac 1 4} \right\rceil, \\
\label{eq: theo: funt, form; j}
\jsoli &= \left\lceil 1 + \frac{\sqrt{\len^2-(\isoli-2)^2a^2}}{b} \right\rceil.
\end{align}
The function $\funt$ is piecewise constant and left-continuous, with unit-height jumps. A jump occurs at $\len$ if and only if $\len = \funl(\tiles)$ for some $\tiles \in \mathbb N$, $\tiles \geq 4$; and then $\funt(\len) = \tiles-1$, $\lim_{h \rightarrow 0+} \funt(\len+h) = \tiles$.
\end{theorem}

@@

The pair $(\isolr, \jsolr)$ can be obtained as a solution of the equation system
\begin{align}
\label{eq: icont jcont 1}
(\isolr-2)^2 a^2 + (\jsolr-2)^2 b^2 &= \len^2, \\
\label{eq: icont jcont 2}
\jsolr = \frac{\isolr a^2}{b^2} - \frac{5a^2}{2b^2} + \frac 3 2.
\end{align}
Expressing these equations in terms of new variables $\isolr-5/2$ and $\jsolr-3/2$, the solutions are easily found to be
\begin{align}
\label{eq: isolr sol}
\isolr &= \frac 5 2 \pm \frac b a \sqrt{\frac{\len^2}{a^2+b^2} - \frac 1 4}, \\
\label{eq: jsolr sol}
\jsolr &= \frac 3 2 \pm \frac a b \sqrt{\frac{\len^2}{a^2+b^2} - \frac 1 4},
\end{align}
where the same sign (positive or negative) should be used in the two equations. This can yield zero, one or two real-valued solution pairs $(\isolr, \jsolr)$, which respectively corresponds to the solid line in Figure~\ref{fig: ijLS_1p35_detail} being exterior, tangent or secant to the circle (the figure depicts the latter case).

A solution pair $(\isolr, \jsolr)$ from \eqref{eq: isolr sol} and \eqref{eq: jsolr sol} is meaningful only if it is real-valued with $\isolr, \jsolr \geq 2$. This solution, if it exists, is always associated with the positive sign in those equations. It is easily seen that $\jsolr \geq 2$ implies $\isolr \geq 2$, and thus it suffices the check the former. Three cases need to be distinguished: there are no real-valued solution pairs $(\isolr, \jsolr)$; there are one or two but none of them has $\jsolr \geq 2$; or there are one or two and and one of them satisfies that inequality. These cases correspond to different ranges of $\len$, as discussed next.

For $\len < \sqrt{a^2+b^2}/2$, \eqref{eq: isolr sol} and \eqref{eq: jsolr sol} are not real-valued. Since $a \geq b$, from the inequality $\len < \sqrt{a^2+b^2}/2$ it follows that $\len < a$. This means that any achievable $(i,j)$ pair, if any, will have $i=2$. Thus in this case $\isoli$ should be set to $2$; and then the maximum achievable $\jsoli$ will be obtained from \eqref{eq: icont jcont 2}.

For $\sqrt{a^2+b^2}/2 \leq \len < (a^2+b^2) / (2a)$, \eqref{eq: isolr sol} and \eqref{eq: jsolr sol} give either two real-valued solutions or one real-valued double solution for $(\isolr, \jsolr)$; but $\jsolr < 2$. This in turn implies, according to \eqref{eq: icont jcont 2}, that $\isolr < (b^2/a^2+5)/2 < 3$ for $a \geq b$. Thus only pairs with $i=2$ are achievable again for $\len$ in this range, and therefore $\isoli$ must also be $2$.

Lastly, for $\len \geq (a^2+b^2) / (2a)$ the expressions \eqref{eq: isolr sol} and \eqref{eq: jsolr sol} with positive sign give $\isolr, \jsolr \geq 2$, and $\isoli$ should be taken as the greatest integer less than $\isolr$, i.e.~$\lceil \isolr \rceil-1$.

The three cases are unified by taking the real part of the positive-sign version of \eqref{eq: isolr sol} and computing $\isoli = \lceil \isolr \rceil-1$, as can be easily checked. This yields \eqref{eq: theo: funt, form; i}. Once $\isoli$ is known, \eqref{eq: theo: funt, form; j} computes $\jsoli$ as the greatest integer such that $(\isoli a,\jsoli b)$ is within the circle with center $(2a,2b)$ and radius $\len$. This ensures that $(\isoli,\jsoli)$ is achievable with lengths less than $\len$.
